---
title: "capstone_project_work"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(leaflet)
library(tigris)
library(dplyr)
library(readr)
library(lubridate)
library(stringr)
library(sf)
library(ggplot2)
library(tidyr)
library(purrr)
library(forcats)
```

# Load in & Wrangle Data

```{r warning=FALSE, message=FALSE}
# read in powerplant location & characteristic data
powerPlants <- read_csv("Data/Power_Plants.csv")
mn_powerplants <- powerPlants %>%
  filter(State == "Minnesota") %>% 
  mutate(fossil_fuel = ifelse(PrimSource %in% c("coal", "petroleum", "natural gas"), "Fossil Fuel", "Renewable"),
         PrimSource = ifelse(PrimSource == "other", "waste heat", PrimSource))  %>% # change 'other' to 'waste heat'
  janitor::clean_names() 
```

Note: air quality only goes 2008-2021

```{r warning=FALSE, message=FALSE,results='hide'}
# read in air quality data
# mn_aq_all_years <- read_csv("Data/Modeled_PM25_Ozone_MN_county_data_allyears.csv") %>% 
#   mutate(county = str_split_i(county, ",", 1),
#          county = str_remove(county, "County"),
#          county = str_trim(county)
#          )
```

Dataset: Form EIA-860

```{r}
# load in dataset containing powerplant first operational date
powerplant_dates <- read_csv("Data/powerplant_data_eia_2024_generator_operable.csv") %>% 
  janitor::clean_names() %>% 
  filter(state == "MN")

# filter to MN and get date of *first* operational generator for a powerplant (disregard dates of generators added later)
powerplant_dates_mn <- powerplant_dates %>% 
  mutate(full_date = make_date(year = operating_year, month = operating_month, day = 1)) %>% 
  group_by(plant_code) %>% 
  summarise(first_op_date = min(full_date))

# add years to mn_powerplants
mn_powerplants <- mn_powerplants %>% left_join(powerplant_dates_mn)

# fill in missing dates:
# Buffalo Sun CSG - active since 2/1/2025
# Oster Sun CSG	- active since 1/1/2025
# Quarry Sun CSG - active since 1/1/2025
# via sunshare website
mn_powerplants <- mn_powerplants %>%
  mutate(first_op_date = case_when(plant_code == 66070 ~ as.Date(mdy("2/1/2025")),
                                   plant_code == 66072 ~ as.Date(mdy("1/1/2025")),
                                   plant_code == 66073 ~ as.Date(mdy("1/1/2025")),
                                   TRUE ~ first_op_date), 
         first_op_year = year(first_op_date)
         ) 

mn_powerplants %>% filter(plant_code == 1912)
```

```{r}

# read in dataset containing powerplant retirement date, if applicable

# powerplant_dates_retired_mn <- read_csv("Data/powerplant_data_eia_egrid_2024_generator_retired.csv") %>%

#   janitor::clean_names() %>%

#   filter(state == "MN")

# if retired utility code not in

# # get date of *last* generator's retirement for a powerplant (disregard dates of generators retired later)

# powerplant_dates_retired_mn <- powerplant_dates_retired_mn %>%

#   mutate(full_retirement_date = make_date(year = retirement_year, month = retirement_month, day = 1)) %>%

#   group_by(plant_code) %>%

#   summarise(last_retire_date = max(full_retirement_date, na.rm = TRUE)) # those with missing dates don't appear in mn_powerplants & can be excluded

#

# # join to mn_powerplants

# mn_powerplants <- mn_powerplants %>% left_join(powerplant_dates_retired_mn) %>% mutate(last_retire_year = year(last_retire_date))

```

```{r warning=FALSE, message=FALSE,results='hide'}

# load spatial/boundary info

mn_counties <- counties(state = "MN", cb = TRUE) %>%

  st_transform(crs = 4326)

mn_tracts <- tracts(state = "MN", cb = TRUE) %>%

  st_transform(crs = 4326)

```

```{r}

# join AQ data to add spatial data

mn_counties_aq <- mn_counties %>%

  left_join(mn_aq_all_years, by = c("NAME" = "county")) %>%

  janitor::clean_names() %>%

  mutate(unit_short = ifelse(unit == "micrograms per cubic meter", "µg/m3", "ppb"))

```

# Visualizations

```{r}

pal_pp <- colorFactor(

  c("black",

    "darkgreen"),

  domain = mn_powerplants$fossil_fuel

)

# Plotting all power plants that have existed in MN

leaflet() %>%

  # putting layer of counties

  addPolygons(

    data = mn_counties,

    fillColor = "lightgray",

    weight = 1,

    color = "darkgray",

    fillOpacity = .6,

    group = "Counties"

  ) %>%

  # putting layer of tracts

  addPolygons(

    data = mn_tracts,

    fillColor = FALSE,

    weight = 0.5,

    color = "gray",

    opacity = 0.5,

    group = "Tracts"

  ) %>%

  # putting circles for each power plant

  addCircleMarkers(

    data = mn_powerplants,

    lng = ~ longitude,

    lat = ~ latitude,

    radius = 0.25,

    color = ~pal_pp(fossil_fuel),

    opacity = 0.35

  ) %>%

  addLegend(

  pal = pal_pp,

  values = mn_powerplants$fossil_fuel,

  title = "Power Plants: Energy Type"

  )

```

This map shows the locations of all powerplants that have been built in Minnesota since 1900, with black points being fossil-fuel powered and green points being renewable-powered. The power plants seem clustered in the southern half of the state, particularly renewable plants, which seem to outnumber fossil-fuel plants.

```{r}

map_mn_aq <- function(year_spec, pollutant_spec){

  # filter df based on user specs

  filtered_mn_aq <- mn_counties_aq %>% filter(year <= year_spec, pollutant == pollutant_spec)

  filtered_mn_powerplants <- mn_powerplants %>% filter(first_op_year <= year_spec, is.na(last_retire_year) | last_retire_year > year_spec)

  unit_spec <- ifelse(pollutant_spec == "Ozone", "ppb", "µg/m3")

    pal_pp <- colorFactor(

    c("black",

      "darkgreen"),

    domain = filtered_mn_powerplants$fossil_fuel

  )

  pal_aqi <- colorNumeric("YlOrRd", domain = filtered_mn_aq$average_concentration)

  # Plotting power plants

  leaflet() %>%

    # putting layer of counties

    addPolygons(

      data = filtered_mn_aq,

      fillColor = ~ pal_aqi(average_concentration),

      weight = 1,

      color = "black",

      fillOpacity = .6,

      group = "Counties"

    ) %>%

    # putting layer of tracts

    addPolygons(

      data = mn_tracts,

      fillColor = FALSE,

      weight = 0.5,

      color = "gray",

      opacity = 0.5,

      group = "Tracts"

    ) %>%

    # putting circles for each power plant

    addCircleMarkers(

      data = filtered_mn_powerplants,

      lng = ~ longitude,

      lat = ~ latitude,

      radius = 0.25,

      color = ~pal_pp(fossil_fuel),

      opacity = 0.35

    ) %>%

    addLegend(

    pal = pal_pp,

    values = filtered_mn_powerplants$fossil_fuel,

    title = "Power Plants: Energy Type"

    )  %>%

    addLegend(

    pal = pal_aqi,

    values = filtered_mn_aq$average_concentration,

    title = str_c(pollutant_spec, " Concentration (", unit_spec, ")")

    )

}

```

```{r}

map_mn_aq(2008, "Ozone")

map_mn_aq(2015, "Ozone")

map_mn_aq(2021, "Ozone")

map_mn_aq(2008, "PM2.5")

map_mn_aq(2010, "PM2.5")

map_mn_aq(2015, "PM2.5")

map_mn_aq(2018, "PM2.5")

map_mn_aq(2021, "PM2.5")

```

This plot allows for a user-specified year and pollutant, to the end of becoming an interactive plot with a slider for year. We can see a lot more renewable plants getting built in later years, but air quality doesn't seem to follow clear patterns over time. Connections between air quality and presence of power plants are not immediately apparent.

```{r}

mn_powerplants %>%

  ggplot(aes(x = first_op_year, fill = fossil_fuel)) +

  geom_bar() +

  facet_wrap(~fossil_fuel) +

  labs(title = "When have different kinds of powerplants been built in Minnesota?",

       x="First Year of Operation",

       y = "Number of Powerplants",

       fill = "Powerplant Type") +

  theme_classic() +

  theme(axis.text.x = element_text(angle = 90, vjust = 1))

# interesting that earliest powerplants are renewable - they're all hydroelectric power

mn_powerplants %>% filter(first_op_year < 1920)

# TO DO: make plot of de-op pps - when do they get decomissioned?

```

The plots show the shift to renewable power around the turn of the century, preceding a huge boom in their construction around 2018. We can see that fossil fuel plant construction has remained fairly constant post-1950, with a lull in the 80s and a spike in the early 2000s. However, the numbers of new fossil fuel plants become nearly insignificant compared to the overwhelming amounts of new renewable plants.

The earliest plants were also renewable ones, making use of exclusively hydroelectric power as shown in the table.

```{r}

mn_powerplants %>%

  ggplot(aes(x = fct_infreq(prim_source))) +

  geom_bar() +

  labs(x = "Energy Source", y = "Number of Powerplants", title = "Minnesota Powerplants' Energy Sources",

      fill = "Powerplant Type") +

  theme_classic() +

  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

mn_powerplants %>%

  group_by(prim_source) %>%

  summarise(prod_by_type = sum(total_mw)) %>%

  mutate(fossil_fuel = ifelse(prim_source %in% c("coal", "petroleum", "natural gas"), "Fossil Fuel", "Renewable")) %>%

  ggplot(aes(x = fct_reorder(prim_source, prod_by_type, .desc = TRUE), y = prod_by_type, fill = fossil_fuel)) +

  geom_bar(stat = 'identity') +

  labs(x = "Energy Source", y = "Megawatts Electricity Produced", title = "Energy produced by Source in MN Powerplants",

       fill = "Powerplant Type") +

  theme_classic() +

  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# TO DOs:

# consider time frame - specific year? this is just showing all that have ever existed

```

Across all plants built in MN, we can see that by far the most (by nearly five times) solar-fueled plants have been built, by numbers alone. There seem to be very few coal plants, although petroleum and natural gas have a notable amount. However, looking at the actual energy produced by each fuel type, natural gas produces the most, followed by wind and coal, suggesting that fossil-fuel plants are either larger and/or more efficient - a single fossil-fuel plant can produce more energy than a renewable one. Solar power produces only about 1800 MWs, less than a third of natural gas's production, despite having over 10x the number of plants.

```{r}

# which counties to look at? ones with new fossil fuel plants in our time range

mn_powerplants %>% filter(fossil_fuel == "Fossil Fuel", first_op_year > 2007) %>% count(county) %>% arrange(desc(n)) %>% head()

```

This table shows which counties have the most new fossil fuel plants built within the time frame of our air quality data (2008-2021), for the sake of analysis.

```{r}

# line chart of county or groups of counties - showing air quality over time; vertical lines when plants were built

#county_spec <- "Dakota"

county_spec <- "Olmsted"

pollutant_spec <- "Ozone"

#pollutant_spec <- "PM2.5"

unit_spec <- ifelse(pollutant_spec == "Ozone", "ppb", "µg/m3")

mn_powerplants_filtered <- mn_powerplants %>%

  filter(county == county_spec, first_op_year >= 2008, first_op_year <= 2021) %>%

  mutate(first_op_date_jittered = first_op_date + ifelse(fossil_fuel == "Fossil Fuel", 7, -7)) # jitter so lines don't cover each other

ggplot() +

  geom_line(data = mn_counties_aqi %>% filter(name == county_spec, pollutant == pollutant_spec) %>%

    mutate(year = as.Date(paste0(year, "-01-01"))), aes(x = year, y = average_concentration)) +

  geom_vline(data = mn_powerplants_filtered, aes(xintercept = first_op_date_jittered, color = fossil_fuel)) +

  scale_color_manual(

    name = "Plant Fuel Type",

    values = c("Renewable" = "aquamarine3", "Fossil Fuel" = "#EE6A50")

  ) +

  labs(title = str_c("Average ", pollutant_spec, " Concentration & Power Plants Built in ", county_spec, " County"),

       y = str_c("Average ", pollutant_spec, " Concentration (", unit_spec, ")"),

       x = "Year") +

  theme_classic()

```

This plot is also designed to be interactive. In this case, there appears little connection between air quality and fossil fuel plants being built.

```{r}

#Overall trends in AQ

mn_counties_aqi %>%

  filter(pollutant == "Ozone") %>%

  ggplot() +

    geom_line(aes(x = year, y = average_concentration, group = name, color = name), show.legend = FALSE) +

  labs(title = "Average Ozone Concentration By County") +

  ylim(0,40) +

  theme_classic()

mn_counties_aqi %>%

  filter(pollutant == "Ozone") %>%

  group_by(year) %>%

  summarise(avg_conc_by_year = mean(average_concentration)) %>%

  ggplot() +

    geom_line(aes(x = year, y = avg_conc_by_year)) +

  labs(title = "Ozone Concentration (Average of All MN Counties)") +

  ylim(0,38) +

  theme_classic()

```

Average ozone concentration does not seems to change too drastically in this time frame; although it dips somewhat around 2016 before rising again. All counties seems to follow the overall trend, with a more distinctive high around 2013 and low around 2015-2016.

```{r}

mn_counties_aqi %>%

  filter(pollutant == "PM2.5") %>%

  ggplot() +

    geom_line(aes(x = year, y = average_concentration, group = name, color = name), show.legend = FALSE) +

  labs(title = "Average PM2.5 Concentration By County") +

  ylim(0, 12) +

  theme_classic()

mn_counties_aqi %>%

  filter(pollutant == "PM2.5") %>%

  group_by(year) %>%

  summarise(avg_conc_by_year = mean(average_concentration)) %>%

  ggplot() +

    geom_line(aes(x = year, y = avg_conc_by_year)) +

  labs(title = "PM2.5 Concentration (Average of All MN Counties)") +

  ylim(0, 10) +

  theme_classic()

# What's causing the 2021 spike?

```

There seems to be more variation in PM2.5, and a distinct pattern shared across counties that suggests larger-scale factors (such as large wildfires) affect each county's air quality more than individual elements like the addition of powerplants. The is a zig-zag pattern between 2014 and 2020, in which PM2.5 jumps and then drops every few years. 2021 measurements seems to be rising back to, or even exceeding pre-2014 levels.

```{r}

# combine the two

#pollutant_spec <- "Ozone"

pollutant_spec <- "PM2.5"

unit_spec <- ifelse(pollutant_spec == "Ozone", "ppb", "µg/m3")

summary_df <- mn_counties_aq %>%

  filter(pollutant == pollutant_spec) %>%

  group_by(name) %>%

  summarize(mean_aqi = mean(average_concentration, na.rm = TRUE)) %>%

  left_join(

    mn_powerplants %>%

      group_by(county) %>%

      count(fossil_fuel) %>%

      filter(fossil_fuel == "Fossil Fuel"),

    by = c("name" = "county")

  ) %>%

  mutate(num_fossil = ifelse(is.na(n), 0, n))

ggplot(summary_df, aes(x = num_fossil, y = mean_aqi)) +

  geom_point(color = "red") +

  geom_smooth(method = 'lm',color = "black") +

  labs(title = str_c("MN Counties' Air Quality vs Number of Fossil Fuel Plants"),

       x = "Number of Fossil Fuel Plants",

       y = str_c("Mean ", pollutant_spec, " Concentration (", unit_spec, ")")) +

  theme_classic()

# renewables

summary_df <- mn_counties_aq %>%

  filter(pollutant == pollutant_spec) %>%

  group_by(name) %>%

  summarize(mean_aqi = mean(average_concentration, na.rm = TRUE)) %>%

  left_join(

    mn_powerplants %>%

      group_by(county) %>%

      count(fossil_fuel) %>%

      filter(fossil_fuel == "Renewable"),

    by = c("name" = "county")

  ) %>%

  mutate(num_fossil = ifelse(is.na(n), 0, n))

ggplot(summary_df, aes(x = num_fossil, y = mean_aqi)) +

  geom_point(color = "turquoise") +

  geom_smooth(color = "black") +

  labs(title = str_c("MN Counties' Air Quality vs Number of Renewable Plants"),

       x = "Number of Renewable Plants",

       y = str_c("Mean ", pollutant_spec, " Concentration (", unit_spec, ")")) +

  theme_classic()

```

A county having more fossil fuel plants appears to be associated with higher PM2.5 concentrations on average. However, having fewer renewable power plants seems associated with better air quality also, which calls this analysis into question, as renewable plants should not be affecting the air quality.

```{r}

summary_df <- mn_counties_aq %>%

  filter(pollutant == pollutant_spec) %>%

  group_by(name) %>%

  summarize(mean_aqi = mean(average_concentration, na.rm = TRUE)) %>%

  left_join(

    mn_powerplants %>%

      group_by(county) %>%

      summarize(has_fossil = any(fossil_fuel == "Fossil Fuel")),

    by = c("name" = "county")

  ) %>%

  mutate(has_fossil = ifelse(is.na(has_fossil), FALSE, has_fossil),

         plant_group = ifelse(has_fossil, "Contains Fossil Fuel Plants", "No Fossil Fuel Plants"))

ggplot(summary_df, aes(x = plant_group, y = mean_aqi, fill = plant_group)) +

  geom_boxplot(alpha = 0.7, outlier.color = "gray40") +

  geom_jitter(width = 0.15, alpha = 0.6) +

  labs(

    title = str_c("Mean ", pollutant_spec, " Concentration by County Type"),

    x = NULL,

    y = str_c("Mean ", pollutant_spec, " Concentration (", unit_spec, ")")

  ) +

  ylim(0,10) +

  scale_fill_manual(values = c("Contains Fossil Fuel Plants" = "#d95f02",

                               "No Fossil Fuel Plants" = "#1b9e77")) +

  theme_classic() +

  theme(legend.position = "none")

mn_powerplants %>% group_by(county) %>% count(fossil_fuel)

```

This plot suggests that counties containing fossil fuel plants have higher concentrations of pm2.5 than those containing only renewable or no powerplants.

```{r}

# look at lagged air quality - what does the aq do the year after a new ff pp installed? 3 years?

mn_powerplants_lagged_aq <- mn_powerplants %>%

  filter(fossil_fuel == "Fossil Fuel",

         first_op_year >= 2008, first_op_year <= 2021) %>%

  mutate(year_after_pp_installed = first_op_year + 1,

         year_3_after_pp_installed = first_op_year + 3) %>%

  left_join(mn_counties_aq %>% select(year, name, pollutant, average_concentration, unit, geometry), by = c("county" = "name", "first_op_year" = "year")) %>%

  rename(average_conc_baseline = average_concentration) %>%

  left_join(mn_counties_aq %>% select(year, name, pollutant, average_concentration, unit, geometry), by = c("county" = "name", "year_after_pp_installed" = "year", "pollutant", "geometry", "unit")) %>%

  rename(average_conc_year1 = average_concentration) %>%

    left_join(mn_counties_aq %>% select(year, name, pollutant, average_concentration, unit, geometry), by = c("county" = "name", "year_3_after_pp_installed" = "year", "pollutant", "geometry", "unit")) %>%

  rename(average_conc_year3 = average_concentration) %>%

  mutate(year1_diff = average_conc_year1 - average_conc_baseline,

         year3_diff = average_conc_year3 - average_conc_baseline)

mn_powerplants_lagged_aq %>%

  group_by(pollutant) %>%

  summarise(mean(year1_diff), mean(year3_diff, na.rm = TRUE)) # have to exclude the year three lag for plant 65405 since that falls in 2022, outside of aq range

```

It appears that air quality improves (pollutant concentration goes down) after fossil fuel plants are added; however, looking at the overall air quality

```{r}

ff_status <- mn_powerplants %>%

  group_by(county) %>%

  summarize(has_fossil = any(fossil_fuel == "Fossil Fuel")) %>%

  mutate(plant_group = ifelse(has_fossil, "Has Fossil Fuel", "Only Renewable/None"))

mn_counties_aq %>%

  filter(pollutant == "PM2.5") %>%

  left_join(ff_status, by = c("name" = "county")) %>%

  mutate(plant_group = ifelse(is.na(plant_group), "Only Renewable/None", plant_group)) %>%

  ggplot(aes(x = year, y = average_concentration, group = name, color = plant_group)) +

    geom_line(alpha = 0.2) + # faint individual counties

    stat_summary(aes(group = plant_group), fun = mean, geom = "line", size = 1.5) + # bold mean trend

    labs(

      title = "Average PM2.5 Concentration by County Type",

      color = "County Plant(s) Type",

      y = "Average PM2.5 Concentration (µg/m3)",

      x = "Year"

    ) +

    ylim(0, 12) +

    scale_color_manual(values = c("Has Fossil Fuel" = "#d95f02",

                                  "Only Renewable/None" = "#1b9e77")) +

    theme_classic()

mn_counties_aq %>%

  filter(pollutant == "Ozone") %>%

  left_join(ff_status, by = c("name" = "county")) %>%

  mutate(avg_conc_jittered = average_concentration + runif(n(), min=0, max=.5)) %>% # jitter so lines don't cover each other

  mutate(plant_group = ifelse(is.na(plant_group), "Only Renewable/None", plant_group)) %>%

  ggplot(aes(x = year, y = avg_conc_jittered, group = name, color = plant_group)) +

    geom_line(alpha = 0.15) + # faint individual counties

    stat_summary(aes(group = plant_group), fun = mean, geom = "line", size = 1) + # bold mean trend

    labs(

      title = "Average Ozone Concentration by County Type",

      color = "County Plant(s) Type",

      y = "Average Ozone Concentration (ppb)",

      x = "Year"

    ) +

    ylim(20, 42) +

    scale_color_manual(values = c("Has Fossil Fuel" = "#d95f02",

                                  "Only Renewable/None" = "lightgreen")) +

    theme_classic()

```

```{r}

# mn_powerplants %>% filter(!is.na(last_retire_date)) %>% group_by(fossil_fuel) %>% count(last_retire_year) %>%

#   ggplot(aes(x=last_retire_year, fill=fossil_fuel, y = n)) +

#   labs(title = "When Have MN Power Plants Been Retired?") +

#   geom_bar(stat="identity") +

#   theme_classic()

#Fossil fuel generators started being retired around 2000, at a pretty constant rate of 1-3 per year.

```

```{r}

# pps and community tool data - from egrid and ACS - pp, emissions, and demographic info

unif_buffers <- read_csv("../../uniform-buffers.csv") %>% janitor::clean_names()

unif_buffers %>% filter(plant_state_abbreviation == "MN")

# plant_id = plant_code in mn_pps

left_join(mn_powerplants, unif_buffers, by = join_by(plant_code == plant_id)) %>% count(is.na(plant_county_name))

# 4 excluded

```

```{r}

# emissions data

NEI <- read_csv("../../2021_NEI_Facility_summary.csv")

NEI %>% filter(state == "MN")

```

```{r}

# new aq data - 2024 only!!!

AirData <- read_csv("../../annual_conc_by_monitor_2024.csv") %>% janitor::clean_names() %>% filter(state_code == 27)

AirData %>%  group_by(state_code, county_code, site_num) %>%

  slice(1) %>% ungroup()

AirData <- AirData %>%

  mutate(popup = paste0(

    "<strong>", local_site_name, "</strong><br/>",

    "Parameter: ", parameter_name, "<br/>",

    #"Sample Duration: ", `Sample Duration`, "<br/>",

    "City: ", city_name, ", ", state_name, "<br/>",

    "Obs Count: ", observation_count))

# only 51 sites - have to pick pollutant. deal with datums

leaflet(AirData %>% filter(parameter_name == "PM2.5 - Local Conditions")) %>%

  addProviderTiles(providers$CartoDB.Positron) %>%

  addCircleMarkers(

    lng = ~longitude, lat = ~latitude,

    radius = 5,

    stroke = FALSE,

    fillOpacity = 0.8,

    popup = ~popup,

    label = ~paste0(local_site_name, " (", parameter_name, ")")

  )  %>%

  # putting circles for each power plant

  addCircleMarkers(

    data = mn_powerplants %>% filter(fossil_fuel == "Fossil Fuel"),

    lng = ~ longitude,

    lat = ~ latitude,

    radius = 0.25,

    color = ~pal_pp(fossil_fuel),

    opacity = 0.35,

    label = ~paste0(plant_name, ", ", utility_name, " (", plant_code, ")")

  )

AirData %>% filter(parameter_name == "PM2.5 - Local Conditions") %>% count(`Site Num`)

```

```{r}

# plants built in 1998 & 2017 - look at aq before and after

mn_powerplants %>% filter(plant_code %in% c(59197, 57966))

# aq monitors in the vicinity - near I35 is closer, should show stronger emissions (likely confounded by highway proxmity)

AirData %>% filter(site_num %in% c("0963", "0962", "1904"),

                   parameter_name == "PM2.5 - Local Conditions",

                   pollutant_standard == "PM25 24-hour 2006",

                   is.na(method_name)

                   )

# really strange that Andersen School has more Pm2.5 - it's right next to a park & hospital; whereas the monitor near the highways & powerplant has the lowest emissions. Don't know how to explain this

# plant in ramsey - 1912 built in 2008, further away from aq monitors but bigger emissions

mn_powerplants %>% filter(plant_code %in% c(55824, 1912))

# aq monitors in the vicinity (likely confounded by highway proximity)

AirData %>% filter(site_num %in% c("0871", "0868"),

                   parameter_name == "PM2.5 - Local Conditions",

                   sample_duration == "24 HOUR",

                   pollutant_standard == "PM25 24-hour 2024", # why are there two? different pocs - two monitors at same site

                   #is.na(method_name)

                   )

# harding is further away, not in city center, should be better aq - true

```

Read in airdata for all years

```{r}
AirData_allyears <- read_csv("Data/airdata_clean.csv")

# get down to 1/site_num/year
# AirData_allyears %>%
#   filter(pollutant_standard == "PM25 24-hour 2012") %>% # 2012 is the most recent standard with the most observations, doesn't change the arith mean
#   count(site_num, year)
# PM2.5 measurements start in 1999. Before that - 1971 TSP (total suspended particulate) was reported; 1987 pm10 reported; 1997 pm2.5


# if multiple POCs, take average
AirData_allyears <- AirData_allyears %>%
  filter(pollutant_standard == "PM25 24-hour 2012") %>% 
    group_by(site_num, parameter_code, pollutant_standard, year, longitude, latitude, local_site_name, county_name, city_name, cbsa_name) %>%
    summarize(avg_pm25 = mean(arithmetic_mean)) %>% 
    mutate(county_name = ifelse(county_name == "Saint Louis", "St Louis", county_name)) %>% 
  ungroup()

AirData_allyears %>% count(year)


```

```{r}

# add legend and color/size scheme 
leaflet(AirData_allyears) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    lng = ~longitude, lat = ~latitude,
    radius = 5,
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste0(round(avg_pm25, digits=1), " µg/m3 - ", local_site_name, " (", county_name, " County)")
  )  %>%
  # putting circles for each power plant
  addCircleMarkers(
    data = mn_powerplants %>% filter(fossil_fuel == "Fossil Fuel"),
    lng = ~ longitude,
    lat = ~ latitude,
    radius = 0.25,
    color = "black",
    opacity = 0.35,
    label = ~paste0(plant_name, ", ", utility_name, " (", plant_code, ")")
  ) 
```

```{r}
# compare monitors with no pps to ones with pps in a spherical buffer - note that some are more likely to actually affect the monitor based on common wind directions

AirData_sf <- st_as_sf(AirData_allyears, coords = c("longitude", "latitude"), crs = 4326)  # WGS84
AirData_sf_proj <- st_transform(AirData_sf, 26915)

mn_pp_sf <- st_as_sf(mn_powerplants, coords = c("longitude", "latitude"), crs = 4326) %>% filter(fossil_fuel == "Fossil Fuel")
mn_pp_proj <- st_transform(mn_pp_sf, 26915)

# Buffer distance: 1 mile = 1609.34 meters
buffer_distance <- 1609.34*3

air_buffers_proj <- st_buffer(AirData_sf_proj, dist = buffer_distance)

within_3miles_of_monitor <- st_within(mn_pp_proj, air_buffers_proj)
mn_pp_sf$near_monitor <- lengths(within_3miles_of_monitor) > 0

# get count of pps within range of an air monitor
pp_within_each_monitor <- st_intersects(air_buffers_proj, mn_pp_proj)
# Add the count as a column
air_buffers_proj$nearby_pp_count <- lengths(pp_within_each_monitor)

air_buffers <- st_transform(air_buffers_proj, 4326) #%>% ungroup()

ggplot() +
  geom_sf(data=air_buffers, aes(fill = nearby_pp_count), alpha = .5) +
  scale_fill_viridis_c(direction = -1) +
  geom_sf(data=mn_pp_sf, size = .5, alpha = .6)


leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  # --- 3-mile buffers around monitors ---
  addPolygons(
    data = air_buffers %>% filter(year == 2015),
    fillColor = ~if_else(nearby_pp_count >0, "red", "blue"),
    fillOpacity = 0.2,
    color = ~if_else(nearby_pp_count > 0, "red", "blue"),
    weight = 1,
    label = ~paste0("Monitor: ", local_site_name, " - ", nearby_pp_count, " plants nearby")
  ) %>%
  # --- Power plants ---
  addCircleMarkers(
    data = mn_pp_sf,
    radius = 2,
    color = "gray",
    fillOpacity = 0.8,
    label = ~paste0("Power Plant: ", plant_name)
  ) 



leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  # --- 3-mile buffers around monitors ---
  addPolygons(
    data = air_buffers %>% filter(year == 2015),
    fillColor = ~if_else(nearby_pp_count > 0, "red", "blue"),
    fillOpacity = 0.2,
    color = ~if_else(nearby_pp_count > 0, "red", "blue"),
    weight = 1,
    label = ~paste0("Monitor: ", local_site_name, " - ", nearby_pp_count, " plants nearby")
  ) %>%
  # --- Power plants ---
  addCircleMarkers(
    data = mn_pp_sf,
    radius = 2,
    color = "gray",
    fillOpacity = 0.8,
    label = ~paste0("Power Plant: ", plant_name)
  ) %>%
  # --- Air monitors ---
  addCircleMarkers(
    data = AirData_sf %>% filter(year == 2015),
    radius = .5,
    color = "purple",
    fill = TRUE,
    fillOpacity = 1
  )

```

```{r}
# compare monitors with no plants to monitors with plants, at various years

calc_avg_pm25 <- function(year_spec){
air_buffers %>% 
  filter(year == year_spec) %>% 
  group_by(nearby_pp_count) %>% 
  summarise(avg_pm25 = mean(avg_pm25)) %>% 
    mutate(year = year_spec)
}

years <- 1999:2024 # incomplete data for 2025; would be inaccurate comparison
library(purrr)
library(ggplot2)
grouped_summ_pm25_allyears <- map(years, calc_avg_pm25) %>% list_rbind()

ggplot(grouped_summ_pm25_allyears) +
  geom_line(aes(x = year, y = avg_pm25, group = as.factor(nearby_pp_count), color = as.factor(nearby_pp_count))) +
  labs(x = "Year", y = "Average PM2.5 Concentration (µg/m3)",
       color = "Number of Plants \nNear Monitor",
       title = "AQ Monitors Near More Plants Report Higher Pollutant Concentrations") +
  theme_classic()

# monitors that are near more power plants tend to have worse aq - that's confounded by the fact that they're in areas like the cities, which are more polluted for many reasons
```

```{r}
# analysis of impact of a new ff pp

plant_monitor_pairs <- st_intersects(air_buffers, mn_pp_sf)

monitor_pp <- tibble(
  site_num = air_buffers$site_num,
  plant_index = pp_within_each_monitor
) %>%
  unnest(plant_index) %>%                             
  mutate(
    plant_id = mn_pp_sf$plant_code[plant_index],
    plant_year = mn_pp_sf$first_op_year[plant_index]
  ) %>%
  select(-plant_index) %>% 
  distinct() # keep only unique rows

# Join to the air-quality time series (one-to-many)
aq_with_pp <- AirData_allyears %>%
  left_join(monitor_pp, by = "site_num")


# ---- 1 year change -----
aq_changes <- aq_with_pp %>%
  filter(!is.na(plant_year)) %>%     # monitors near a plant
  mutate(period = case_when(
    year == plant_year - 1 ~ "before",
    year == plant_year + 1 ~ "after",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period))

pair_changes <- aq_changes %>%
  group_by(site_num, plant_id) %>%
  summarize(
    before = avg_pm25[period == "before"],
    after  = avg_pm25[period == "after"],
    change = after - before
  )

ggplot(pair_changes %>% pivot_longer(3:4, names_to = "period", values_to = "avg_pm25_annual")) +
  geom_point(aes(x = fct_relevel(period, c("before", "after")),  y = avg_pm25_annual, group = site_num, color = site_num),  show.legend = FALSE) +
  geom_line(aes(x = fct_relevel(period, c("before", "after")),  y = avg_pm25_annual, group = site_num, color = site_num),  show.legend = FALSE) +
  theme_minimal() +
  labs(title = "PM2.5 Concentration from Air Monitors Near New Plants",
       x = "Year (Relative to Plant Beginning Operations)",
       y = "PM2.5 Concentration (µg/m3)") 

# leaflet() %>%
#   addProviderTiles("CartoDB.Positron") %>%
#   # --- 3-mile buffers around monitors ---
#   addPolygons(
#     data = air_buffers %>% filter(site_num %in% pair_changes$site_num, year == 2002),
#     fillColor = "blue",
#     fillOpacity = 0.2,
#     color = "blue",
#     weight = 1,
#     label = ~paste0("Monitor: ", local_site_name)
#   ) %>%
#   # --- Power plants ---
#   addCircleMarkers(
#     data = mn_pp_sf %>% filter(plant_code %in% pair_changes$plant_id),
#     radius = 2,
#     color = "gray",
#     fillOpacity = 0.8,
#     label = ~paste0("Power Plant: ", plant_name)
#   )


# ---- 3 year change -----
aq_changes_3yr <- aq_with_pp %>%
  filter(!is.na(plant_year)) %>%     # monitors near a plant
  mutate(period = case_when(
    year == plant_year - 1 ~ "before",
    year == plant_year + 3 ~ "after",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period))

pair_changes_3y <- aq_changes_3yr %>%
  group_by(site_num, plant_id) %>%
  summarize(
    before = avg_pm25[period == "before"],
    after  = avg_pm25[period == "after"],
    change = after - before
  )


air_buffers %>% filter(site_num %in% c("0962", "0963"))
```

```{r}
# could maybe do something similar to what I did before with monitors and new pps in twin cities

# mn_powerplants_filtered <- mn_powerplants %>% 
#   filter(county %in% c("Ramsey", "Hennepin"), first_op_year >= 1999, first_op_year <= 2024) %>% 
#   mutate(first_op_date_jittered = first_op_date + ifelse(fossil_fuel == "Fossil Fuel", 15, -15)) # jitter so lines don't cover each other
# 
# ggplot(data = AirData_allyears %>% filter(county_name %in% c("Ramsey", "Hennepin")) %>% mutate(year = as.Date(paste0(year, "-01-01"))) %>% 
#          group_by(year) %>% summarise(avg_pm25 = mean(arithmetic_mean)), # average across sites
#     aes(x = year, y = avg_pm25)) +
#   geom_line() +
#   geom_point() +
#   geom_vline(data = mn_powerplants_filtered, aes(xintercept = first_op_date_jittered, color = fossil_fuel), size = .7) + 
#   scale_color_manual(
#     name = "Plant Fuel Type",
#     values = c("Renewable" = "aquamarine3", "Fossil Fuel" = "#EE6A50")
#   ) +
#   labs(title = str_c("Average PM2.5 Concentration & Power Plants Built in the Twin Cities"), 
#        y = str_c("Average PM2.5 Concentration (µg/m3)"),
#        x = "Year") +
#   theme_classic()
```

```{r}
# among counties with aq monitors, compare those that have ff pps - less sensitive version of the above

ff_status <- mn_powerplants %>%
  group_by(county) %>%
  summarize(has_fossil = any(fossil_fuel == "Fossil Fuel")) %>%
  mutate(plant_group = ifelse(has_fossil, "Has Fossil Fuel", "Only Renewable/None"))


AirData_allyears %>%
  left_join(ff_status, by = c("county_name" = "county")) %>% 
  group_by(year, county_name, plant_group) %>% 
  summarize(avg_pm25_grouped = mean(avg_pm25)) %>% 
  mutate(plant_group = ifelse(is.na(plant_group), "Only Renewable/None", plant_group)) %>%
  ggplot(aes(x = year, y = avg_pm25_grouped, group = county_name, color = plant_group)) +
    geom_line(alpha = 0.4) + #  individual counties
    stat_summary(aes(group = plant_group), fun = mean, geom = "line", size = 1.5) + # mean trend
    labs(
      title = "Average PM2.5 Concentration by County Type",
      color = "County Plant(s) Type",
      y = "Average PM2.5 Concentration (µg/m3)",
      x = "Year"
    ) +
    ylim(0, 12) +
    scale_color_manual(values = c("Has Fossil Fuel" = "#d95f02",
                                  "Only Renewable/None" = "#1b9e77")) +
    theme_classic()

# this has so many asterisks 
```

```{r}
# fit a naive model

AirData_allyears$nearby_pp_count <- lengths(pp_within_each_monitor)
# incorporate cbsa
metro_stat_areas <- c("Minneapolis-St. Paul-Bloomington, MN-WI", "Duluth, MN-WI", "St. Cloud, MN", "	Mankato-North Mankato, MN", "Rochester, MN")
AirData_allyears <- AirData_allyears %>% mutate(in_metro_area = cbsa_name %in% metro_stat_areas) 

lm(avg_pm25 ~ latitude + year + nearby_pp_count + in_metro_area, data = AirData_allyears) %>% summary()
# holding all else constant, power plants within 3 miles of a power plant have higher pm2.5 concentrations on average
# pm2.5 decreasing over time
# in metro area not significant
#wildfire smoke, city traffic - can't account for
```

```{r}
# incorporate spatial correlation...

# establish neighborhood structure ???  no - it's point data
```

```{r}
# time series analysis 

# visualize
ggplot(data = AirData_allyears %>% group_by(year) %>% summarize(avg_avg_pm25 = mean(avg_pm25)), aes(y = avg_avg_pm25, x = year)) +
  geom_line() +
  theme_classic()

ggplot(data = AirData_allyears, aes(y = avg_pm25, x = year, group = site_num)) +
  geom_line() +
  theme_classic()

# detrend & decompose
AirData_allyears_yravg <- AirData_allyears %>% group_by(year) %>% summarize(avg_avg_pm25 = mean(avg_pm25))
AirData_allyears_ts <- ts(AirData_allyears_yravg$avg_avg_pm25, start = 1999, frequency = 1) 

# MA
fltr <- rep(1/3, 3) #vector of weights for moving average filter - 3 year avg
AirData_allyears_yravg <- AirData_allyears_yravg %>%
  mutate(pm25_trend_MA = as.numeric(stats::filter(AirData_allyears_ts, filter = fltr, method = 'convo', sides = 2)))

# loess
AirData_allyears_yravg <- AirData_allyears_yravg %>%
  mutate(pm25_trend_loess = predict(loess(avg_avg_pm25 ~ year, data = AirData_allyears_yravg, span = .3)))

# splines 
library(splines)
AirData_allyears_yravg <- AirData_allyears_yravg %>%
  mutate(pm25_trend_spline = predict(lm(avg_avg_pm25 ~ bs(year, knots=c(2004, 2016, 2017, 2023),degree=2), data = AirData_allyears_yravg)))

AirData_allyears_yravg  %>% 
  ggplot(aes(x = year, y = avg_avg_pm25)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y = pm25_trend_MA), color = 'red') +
  geom_line(aes(y = pm25_trend_loess), color = 'blue') +
  geom_line(aes(y = pm25_trend_spline), color = 'green') +
  theme_classic()


# plot residuals
AirData_allyears_yravg <- AirData_allyears_yravg %>%
  mutate(detrend = avg_avg_pm25 - pm25_trend_loess)

AirData_allyears_yravg  %>% 
  ggplot(aes(x = year, y = detrend)) +
  geom_point() +
  geom_line() +
  geom_smooth(se=F) +
  geom_hline(yintercept = 0) +
  theme_classic()
# residuals centered at mean zero, don't seem super correlated 

library(astsa)
acf2(AirData_allyears_yravg$detrend)
# ACF seems to drop after lag 1; maybeee decaying
# PACF possibly decaying; or else dropping after lag 2

# candidate models: 
# ACF drops at 1, PACF decays: MA(1)
sarima(AirData_allyears_yravg$detrend, p = 0, d = 0, q = 1)
# eh

# ACF decays, PACF drops at 2: AR(2) 
sarima(AirData_allyears_yravg$detrend, p = 2, d = 0, q = 0)
# this is muchhh better - higher LB stats and lower residuals

# ARMA(2,1)
sarima(AirData_allyears_yravg$detrend, p = 2, d = 0, q = 1)
# this is the best 
```

```{r}
# Incorp into model


```

```{r}
# make map of power plant size & ff vs not & add context to pp section

library(scales)
leaflet() %>%
      # addTiles() %>%
      addProviderTiles("CartoDB.Positron") %>%
      setView(lng = -94.6, lat = 46.4, zoom = 6) %>%
      addCircleMarkers(data = mn_powerplants, 
                       color = ~if_else(fossil_fuel == "Fossil Fuel", "#d95f02", "#1b9e77"), 
                       radius = ~rescale(total_mw, to = c(1, 16)),
                       #stroke = FALSE,
                       label = ~paste0(plant_name, " (", prim_source, " - ", total_mw, " megawatt(s))")
                       ) %>%
      addLegend(
        position = "bottomright",
        title = "Power Plants by Size",
        colors = c("#d95f02", "#1b9e77"),
        labels = c(
          "Fossil Fuel",
          "Renewable"
        )
      )

```

```{r}
# choose the specific power plant
target_pp <-  st_as_sf(mn_powerplants, coords = c("longitude", "latitude"), crs = 4326) %>% 
   st_transform(26915) %>% 
  filter(str_detect(plant_name, "Covanta"))

# find which air buffers intersect this plant's location
hits <- st_intersects(air_buffers_proj, target_pp)

# keep only buffers that intersect
air_buffers_near_target <- air_buffers_proj %>% filter(lengths(hits) > 0)

# if want WGS84 again 
air_buffers_near_target <- st_transform(air_buffers_near_target, 4326)


herc_monitor_avgs <- air_buffers_near_target %>% 
  group_by(year) %>% 
  summarize(pm25_avg_by_year = mean(avg_pm25)) %>% 
  mutate(id = "Near HERC")

all_monitor_avgs <- air_buffers %>% 
  group_by(year) %>% 
  summarize(pm25_avg_by_year = mean(avg_pm25)) %>% 
  mutate(id = "Whole State")

all_avgs <- rbind(all_monitor_avgs, herc_monitor_avgs)

# diff between avg of all air monitors in the state and air monitors just near the plant
ggplot(all_avgs) +
  geom_line(aes(y=pm25_avg_by_year, x=year, color = id)) +
  scale_color_manual(values = c("Near HERC" = "red", "Whole State" = "black")) +
  labs(title = "PM2.5 Concentrations Are Consistently Higher near the HERC",
       y = "Annual Average PM2.5 Concentration (µg/m3)",
       x = "Year",
       color = "Location of Monitor") +
  theme_classic()

```


# AI Statement

I used ChatGPT to help create the more complicated plots at the end, as well as to help me debug some leaflet errors. I also brainstormed new methods of analysis when my original data fell through and used it to debug various issues throughout.
